Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-09-03 paper/pool_model.tex
Add citations to introduction and correct some mistakes

Bacterial consortia are highly complex systems which consist of multiple distinct cell types.
The underlying processes are interconnected with each other and occur on all scales, ranging from
population-based dynamics of colony growth over molecular processes such as metabolism to gene
expression.
Bacterial communities are ubiquitous and even able to survive under extreme conditions, making them
virtually omnipresent around any living system.
They play an especially important role within the context of food safety and production
being a key contributor to the spoilage of food products as well as fermentation processes.
Furthermore effects such as biofilm formation can hinder drug treatment by shielding the bacterial
colony from antibiotic substances or immune response of the organism.
The natural bacetrial processes are a crucial part of our planets ecosystem, being responsible for
tasks such as nutrient cycling, decomposition of organic matter and oxygen production.
It is of great interest to understand the dynamics of these various processes.

@@

\noindent
In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biologial phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth.
Additionally, bacteria are able to cooperate e.g. via complementary metabolic processes or inhibit
each other e.g. by means of waste production as a metabilic byproduct.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony.
Bacteria have thus devised various survival strategies to ensure a sufficient access to nutrients.
Effects on a single-cell level such as adhesion, migration and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling e.g. in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial community.

@@

\noindent
The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under contant
environmental conditions.
Widely used examples include the modified Gompertz model~\cite{Gompertz1825-wi}, Baranyi-Roberts
model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994} and MCKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\cite{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\cite{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicitiy of usage.
However due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.

@@

\noindent
In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states i.e. pools.
Our model also allows us to capture interactions between these bacterials states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spacial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we asses how spatial heterogeneity can impact the shape of the growth curves.

@@

For homogeneous densities it is equivalent to consider the total amount, i.e., multiplying $L$ and $G$ by the volume.

@@

    \dot{G_A} = \lambda_A R L_A + \alpha_A R G_A\\
    \dot{L_B} = -\lambda_B R L_B\\
    \dot{G_B} = \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B

@@

where $\delta(x-x_\text{cell})$ is the dirac delta-distribution.

@@

Furthermore, we let cells produce an inhibitor with production rate $b$ which inhibits the cells ability to uptake and thus use the resource by a factor of $1+\nu_I [I]$

@@

Thus we introduce an additional rule that splits apart the cell whenever a certain volume threshold is reached.

@@

In addition to modeling division events, the cells have are subject to the lag-phase which delays their ability to proliferate.

@@

We are thus able to derive the \ac{ode} pool model from its \ac{abm} version in the homogenenous case (see supplementary material~\ref{subsec:supplement-deriving-pool-odes}).

@@

Table~\ref{tab:spatial_limit_init} shows a comparison of paramters of two simulations for the \ac{ode} and \ac{abm} model which have been mapped onto each other.

@@

        Darker colours indicate that less nutrients are present.

>>>>

Bacterial consortia are highly complex systems which consist of multiple distinct cell types.
The underlying processes are interconnected with each other and occur on all scales, ranging from
population-based dynamics of colony growth over molecular processes such as metabolism to gene
expression.
Bacterial communities are ubiquitous and even able to survive under extreme conditions, making them
virtually omnipresent around any living system.
They play an especially important role within the context of food safety and production
being a key contributor to the spoilage of food products as well as fermentation processes.
Furthermore, effects such as biofilm formation can hinder drug treatment by shielding the bacterial
colony from antibiotic substances or immune response of the organism.
The natural bacterial processes are a crucial part of our planet's ecosystem, being responsible for
tasks such as nutrient cycling, decomposition of organic matter and oxygen production.
It is of great interest to understand the dynamics of these various processes.

@@

\noindent
In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\cite{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\cite{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate e.g. via complementary metabolic processes or inhibit
each other e.g. by means of waste production as a metabolic byproduct~\cite{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\cite{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\cite{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling e.g. in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\cite{ng_bacterial_2009}.

@@

\noindent
The first step in the construction of any mathematical model is to determine the required level of
detail with respect to the description scale and consequently which effects to include.
The bacterial growth curves let us describe the dynamics of the overall abundance of various
species.
The field of predictive microbiology has developed a variety of ODE-based models to capture
bacterial growth dynamics~\cite{perez-rodriguez_predictive_2012}.
Various \acp{ode} \textit{primary models} describe the bacterial growth curve under constant
environmental conditions.
Widely used examples include the modified Gompertz model~\cite{Gompertz1825-wi}, Baranyi-Roberts
model~\cite{baranyi_modeling_1993, baranyi_dynamic_1994} and McKellar two-compartment model~\cite{buchanan_when_1997} which distinguishes between growing and non-growing bacteria.
To take into account changing environmental conditions, so-called \textit{secondary models}
formulate the model parameters as a function of external quantities such as temperature, pH or
humidity.
These can be simple polynomial models, square root-type models~\cite{ratkowsky_relationship_1982} or
more complex relations such as the cardinal parameter model~\cite{zwietering_decision_1992}.
The advantage of the \ac{ode} formulation lies in the predictive power and simplicity of usage.
However, due to their coarse-grained nature, they lack the ability to describe heterogeneity and
other spatial effects.

@@

\noindent
In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states i.e. pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity can impact the shape of the growth curves.

@@

For homogenous densities it is equivalent to consider the total amount, i.e., multiplying $L$ and $G$ by the volume.

@@

    \dot{G_B} &= \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B

@@

where $\delta(x-x_\text{cell})$ is the Dirac delta-distribution.

@@

Furthermore, we let cells produce an inhibitor with production rate $b$ which inhibits the cell ability to uptake and thus use the resource by a factor of $1+\nu_I [I]$

@@

Thus, we introduce an additional rule that splits apart the cell whenever a certain volume threshold is reached.

@@

In addition to modeling division events, the cells are subject to the lag-phase which delays their ability to proliferate.

@@

We are thus able to derive the \ac{ode} pool model from its \ac{abm} version in the homogenous case (see supplementary material~\ref{subsec:supplement-deriving-pool-odes}).

@@

Table~\ref{tab:spatial_limit_init} shows a comparison of parameters of two simulations for the \ac{ode} and \ac{abm} model which have been mapped onto each other.

@@

        Darker colors indicate that less nutrients are present.
